---
title: "Leveraging Protein Turnover for Drug Discovery: figures"
author: "Jitao david Zhang"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document:
    df_print: paged
    theme: spacelab
    mathjax: default
    code_folding: hide
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
editor_options: 
  chunk_output_type: inline
params:
  echo: yes
  relative: FALSE
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.height=6, fig.width=6)
library(ribiosUtils)
library(ribiosIO)
library(ribiosPlot)
library(openxlsx)
library(tidyverse)
library(ggplot2)
library(emmeans)
library(knitr)
library(rmarkdown)
library(kableExtra)
library(zoo)
library(Vennerable)
library(ggalluvial)
theme_set(theme_light(base_size=13))
```

## Definition of protein turnover

Assuming that the protein follows zero-order synthesis with constant production rate $k_s$, and first-order degradation with rate $k_d$. The ordinary differential equation describing protein concentration changes with regard to $t$ is

$$ \frac{dP}{dt} = k_s - k_dP(t) $$.

The closed form solution, assuming that $ P(0) = 0$, is

$$ P(t) = \frac{k_s}{k_d}(1-e^{-k_dt}) $$
Here the assumption is that protein is synthesized and degraded with a constant synthesis rate and a constant degradation rate following exponential decay. The constant exponential-decay rate suggests that proteins, old or new, have the same probability at any time to be degraded. In particular, even newly synthesized proteins are degraded, so that the accumulation of newly synthesized protein is not linear (as a simple zero-order synthesis rate would suggest), but rather sigmoidal (due to the decay term).

McShane et al. (2016) showed that this model is probably valid for about 90% of the proteins, at least in mouse fibroblasts and human retinal pigmented epithelial cell lines: for the rest 10% proteins show non-exponential decay, i.e. they are less stable in the first hours in their life, and becomes more stable later.

Assuming the zero-order synthesis and first-order degradation model, and assuming that the protein abundance does not change during the period during which we investigate its turnover, then during this time interval the total synthesis equals the total degradation, and the half-life, which is defined by the time when half of the protein species consists of old protein while the other half consists of newly synthesized molecules after the labeling, can be solved analytically:

$$ t_{1/2} = \frac{ln(2)}{k_d} $$

The faster the degradation rate $k_d$, the shorter the turnover.

At the same time, according to the model, the homeostasis concentration $P$ is given as 

$$ P = \frac{k_s}{k_d} $$.

Therefore,

$$ k_s = \frac{ln(2) P}{t_{1/2}} $$.

This means that $k_d$ and $k_s$ can be deduced from the synthesis/degradation curves of the labeling experiment if the model assumptions hold.


```{r homeostasis_prep}
# Define parameters
k_s <- 8     # Zero-order synthesis rate
k_d <- 0.16    # First-order degradation rate
p_homeo <- k_s/k_d # asymptotic steady-state (homeostasis) level
time <- seq(0, 35, length.out = 1000)
spike_time <- 10

## newly 
new_prot_func <- function(t) {
  eff_time <- t - spike_time
  res <- k_s/k_d * (1 - exp(-k_d * eff_time))
  res[eff_time < 0] <- NA
  return(res)
}
old_prot_func <- function(t) {
  eff_time <- t - spike_time
  res <- p_homeo * exp(-k_d * eff_time)
  res[eff_time < 0] <- p_homeo
  return(res)
}
new_prot <- new_prot_func(time)
old_prot <- old_prot_func(time)
halflife_t_ind <- which.min(abs(new_prot - old_prot))
halflife_t <- time[halflife_t_ind]
halflife_new_prot <- new_prot_func(halflife_t)
halflife_old_prot <- old_prot_func(halflife_t)
halflife_avg_prot <- (halflife_new_prot + halflife_old_prot)/2 ## should be exactly half of k_s/k_d

prot_df <- data.frame(time=time,
                      old=old_prot,
                      new=new_prot)
new_prot_col <- "#1a9641"
old_prot_col <- "#fdae61"
total_prot_col <- "#2c7bb6"
```

```{r homeostasis_plot, fig.height=3, fig.width=5}
homeostasis <- ggplot(prot_df, aes(x=time, y=new)) +
  geom_hline(yintercept=0, col="darkgray") +
  geom_line(lwd=1.5, col=new_prot_col) +
  geom_line(aes(x=time, y=old), col=old_prot_col, lwd=1.5) +
  geom_hline(yintercept = p_homeo, col=total_prot_col, lwd=1.5) +
  geom_vline(xintercept = spike_time, col="darkgray") +
  annotate("text", x=spike_time*0.95, y=p_homeo*0.9, label="Labeling",
           vjust=1, hjust=1, size=6, col="darkgray") +
  annotate("text", x=17, y=p_homeo*0.97, col=total_prot_col,
           label="Total", vjust=1, hjust=1, size=6) +
  annotate("text", x=30, y=new_prot_func(20), col=new_prot_col,
           label="New protein", size=6, hjust=1, vjust=1) +
  annotate("text", x=30, y=old_prot_func(20), col=old_prot_col,
           label="Old protein", size=6, hjust=1, vjust=0) +
  annotate("segment", x=halflife_t, xend=halflife_t,
           y=0, yend=halflife_avg_prot, lty=2, col="black") +
  annotate("segment", x=spike_time, xend=halflife_t,
           y=halflife_avg_prot, yend=halflife_avg_prot, lty=2, col="black") +
  annotate("text", x=halflife_t, y=3, hjust=0, size=6,
           label="paste(\"Half-life \", t[1/2])", parse = TRUE) +
  annotate("segment", x=spike_time, xend=halflife_t,
           y=3, yend=3, col="black",
           arrow=arrow(type="closed", ends="both", length=unit(0.03, "npc"))) +
  scale_x_continuous(expand=expansion(mult=0, add=0),
                     limits=range(time),
                     labels=function(x) x-spike_time) +
  scale_y_continuous(limits=c(0, 55), expand=expansion(mult=0, add=0)) +
  xlab("Time [A.U.]") +
  ylab("Protein concentration [A.U.]")
print(homeostasis)
```


A similar concept describing protein turnover is the mean lifetime $\tau$, which is defined by the average length of time a protein species persists. Mathematically, it can be shown that it is the time required for a protein to be reduced to $1/e$ of its initial quantity. The relationship between mean lifetime, degradation rate, and half-life can be expressed as 

$$ \tau = \frac{1}{k_d} = \frac{t_{1/2}}{ln(2)} $$.

Using mean life time simplifies the expression of the relationship between the synthesis rate and the protein abundance

$$ P = k_s \tau $$,

which is also known as [Little's Law](https://en.wikipedia.org/wiki/Little%27s_law) in mathematics.

In the review, when not otherwise indicated, we use protein half-life to express protein's turnover.

# Disclaimer

Through out the code I have used large-language models to generate part of the code. I have checked their accuracy and edited them if necessary.

# Session information

```{r sessionInfo}
sessionInfo()
```